---
title: "class10_halloween_mini_project"
author: "Anita Wang"
date: "10/28/2021"
output: html_document
---
#**1. Importing candy data**

```{r}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy <- read.csv(url, row.names=1)
head(candy, n=5)
```

*What is in the dataset?*

>**Q1. How many different candy types are in this dataset?**

```{r}
nrow(candy)
```

There are 85 different candy types in this dataset

>**Q2. How many fruity candy types are in the dataset?**

```{r}
sum(candy$fruity)
```

There are 38 fruity candy types in the dataset

#**2. What is your favorate candy?**

We can find the winpercent value for Twix by using its name to access the corresponding row of the dataset. This is because the dataset has each candy name as rownames (recall that we set this when we imported the original CSV file). For example the code for Twix is:

```{r}
candy["Twix", ]$winpercent
```

>**Q3. What is your favorite candy in the dataset and what is it’s winpercent value?**

```{r}
candy["Starburst", ]$winpercent
```

Starbursts are my favorite candy. The winpercent value for Starbursts is 67.03763 percent.

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

The winpercent value for "Kit Kat" is 76.7686 percent.

>**Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?**

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent

```

The winpercent value for Toosie Rolls is 49.6535 percent. 

```{r}
#Install skim() function in the skimr package that can help give you a quick overview of a given dataset
#install.packages("skimr")

library("skimr")
skim(candy)
```

>**Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?**

Yes.

>**Q7. What do you think a zero and one represent for the candy$chocolate column?**

1 = TRUE that the candy is a chocolate candy
0 = FALSE that the candy is a chocolate candy


A good place to start any exploratory analysis is with a histogram. You can do this most easily with the base R function hist(). Alternatively, you can use ggplot() with geom_hist(). Either works well in this case and (as always) its your choice.

>**Q8. Plot a histogram of winpercent values**

```{r}
hist(candy$winpercent)
```

>**Q9. Is the distribution of winpercent values symmetrical?**

No

>**Q10. Is the center of the distribution above or below 50%?**

Below

>**Q11. On average is chocolate candy higher or lower ranked than fruit candy?**

First need to find all the chocolate candy rows in the candy dataset

Use as.logical() to change 1's and 0's to TRUE's and False's

```{r}
inds <- as.logical(candy$chocolate)
chocolate <- candy[inds,]$winpercent

inds.fruit <- as.logical(candy$fruity)
fruity <- candy[inds.fruit,]$winpercent

mean(chocolate)
mean(fruity)
```

On average, chocolate candy is higher ranked than fruit candy.

>**Q12. Is this difference statistically significant?**

Use t.test() to run a t-test

```{r}
t.test(chocolate, fruity)
```

p-value = 2.871e-08. Yes, the difference is statistically significant.

#3. **Overall Candy Rankings**

Let’s use the base R order() function together with head() to sort the whole dataset by winpercent. Or if you have been getting into the tidyverse and the dplyr package you can use the arrange() function together with head() to do the same thing and answer the following questions:

>**Q13. What are the five least liked candy types in this set?**

```{r}
head(candy[order(candy$winpercent),], n=5)
```

Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters 

>**Q14. What are the top 5 all time favorite candy types out of this set?**

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE),], n=5)
```

Reese's Peanut Butter cups, Reese's Miniatures, Twix, Kit Kats, and Snickers	

We can also make a barplot to visualize the overall rankings. We will use an iterative approach to building a useful visulization by getting a rough starting plot and then refining and adding useful details in a stepwise process.

>**Q15. Make a first barplot of candy ranking based on winpercent values.**

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

>**Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?**

Improve this plot by reordering the winpercent value:

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

*Time to add some useful color*

Let’s setup a color vector (that signifies candy type) that we can then use for some future plots. We start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values:

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "red"
```

Now plot with color:

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```

>**Q17. What is the worst ranked chocolate candy?**

Nik L nip

>**Q18. What is the best ranked fruity candy?**

Reeses Peanut Butter Cups 


#**4. Taking a look at pricepercent**

Make a plot of winpercent vs the pricepercent variable. The pricepercent variable records the percentile rank of the candy’s price against all the other candies in the dataset. Lower vales are less expensive and high values more expensive

```{r}
library(ggrepel)

# How about a plot of price vs win

#You can improve the clarity of the labels by 
#using the **ggrepl** package and 
#the `geom_text_repel()` function 

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

>**Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?**

```{r}
ord <- order(candy$pricepercent)
head( candy[ord,c(11,12)], n=5 )
```
Tootsie Roll Midgies	

>**Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?**

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. Nik L Nip is the least popular. 

#**5. Exploring the correlation structure**

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Being fruity and being chocolate

>Q23. Similarly, what two variables are most positively correlated?

Chocolate bars and higher pricing for chocolate candies (high winpercent value)


##No need to submit a labreport for this lab. Just push it to gitgub. This lab was just for fun

#6. **Principal Component Analysis**

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
pca$rotation[,1]
```

Now, plot our main PCA score plot of PC1 vs PC2:















