---
title: 'Section 4: Population Scale Analysis [HOMEWORK]'
author: 'Anita Wang (PID: A15567878)'
date: "12/5/2021"
output:
  pdf_document: default
  html_document: default
---
>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. *Hint: The read.table(),summary() and boxplot() functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein. Note that you can find the medium value from saving the output of the boxplot() function to an R object and examining this object.There is also the medium() and summary() function that you can use to check your understanding.*

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

**Sample size for each genotype:**

A/A A/G G/G 
108 233 121 

```{r}
genotype <- expr$geno
express <- expr$exp

med <- boxplot(express~genotype)
med
```

**Corresponding median expression levels for each genotype:**

A/A      A/G      G/G 
31.24847 25.06486 20.07363


Let's make a boxplot

```{r}
library(ggplot2)

```

```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```


>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need to be as polished as this one.

- From the boxplot, you can infer that the SNP does effect the expression of ORMDL3. Having a G/G in this location is associated with having a reduced expression of ORMDL3.
