---
title: 'Class 17: Vaccination rate mini project'
author: 'Anita Wang (PID: A15567878)'
date: "11/23/2021"
output: html_document
---

#Getting Started

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```


>Q1. What column details the total number of people fully vaccinated? 

- The 9th column: "persons_fully_vaccinated"

>Q2. What column details the Zip code tabulation area? 

- The 2nd column

>Q3. What is the earliest date in this dataset? 

```{r}
head(vax$as_of_date)
```

- 2021-01-05 (January 5th, 2021)

>Q4. What is the latest date in this dataset? 

```{r}
tail(vax$as_of_date)
```

- 2021-11-16 (November 16th, 2021)

As we have done previously, let’s call the skim() function from the skimr package to get a quick overview of this dataset:


```{r}
skimr::skim(vax)
```

>Q5. How many numeric columns are in this dataset? 

- 9 columns

>Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

- 8256 NA values

>Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
per_vax <- sum(is.na(vax$persons_fully_vaccinated)) / sum(vax$persons_fully_vaccinated, na.rm=TRUE)

signif((per_vax *100), 2)

```

- 0.0012%

>Q8. [Optional]: Why might this data be missing?

- The areas don't report their vaccination. For example, the military has their own health systems and does not participate in this record keeping.

#Working with dates

##Ensure that the date column is useful

We will use the **lubridate** package to make life a lot easier when dealing with date and times 

```{r}
#install lubridate
#install.packages("lubridate")

library(lubridate)

```


```{r}
today()
```

Here we make our ~as-of-date` column lubridate format

```{r}
# Speciffy that we are using the Year-mont-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

Now we can do math with dates. For example: How many days have passed since the first vaccination reported in this dataset?

```{r}
today() - vax$as_of_date[1]

```

Using the last and the first date value we can now determine how many days the dataset span?

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]

```

>Q9. How many days have passed since the last update of the dataset? 

```{r}
today() - vax$as_of_date[nrow(vax)]
```

- 7 days

>Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
length(unique(vax$as_of_date))
```

-There are 46 unique dates in the dataset 

#Working with ZIP codes

In R we can use the zipcodeR package to make working with these codes easier.

```{r}
#install.packages("zipcodeR")
library(zipcodeR)
geocode_zip('92037')
```

Calculate the distance between the centroids of any two ZIP codes in miles, e.g.

```{r}
zip_distance('92037','92109')

```

More usefully, we can pull census data about ZIP code areas (including median household income etc.). For example:

```{r}
reverse_zipcode(c('92037', "92109") )
```

We can use this reverse_zipcode() to pull census data later on for any or all ZIP code areas we might be interested in.

```{r}
# Pull data for all ZIP codes in the dataset
#zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

#Focus on the San Diego area

Let’s now focus in on the San Diego County area by restricting ourselves first to vax$county == "San Diego" entries. We have two main choices on how to do this. The first using base R the second using the dplyr package:

Using base R, he code would look like this: 

```{r}
# Subset to San Diego county only areas
sd <- vax[ vax$county == "San Diego" , ]
```

Using dplyr the code would look like this:

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

Using dplyr is often more convenient when we are subsetting across multiple criteria (more advanced subsetting) - for example all San Diego county areas with a population of over 10,000.

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```


>Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
length(unique(sd$zip_code_tabulation_area))
```

- There are 107 distinct zip codes 

>Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
which.max(sd$age12_plus_population)

```

```{r}
sd[23,]
```

- zip code: 92154


Using dplyr select all San Diego “county” entries on “as_of_date” “2021-11-16” and use this for the following questions.

>Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”? 

```{r}
library(dplyr)

sd.now <- filter(sd, as_of_date ==  "2021-11-09")
(mean(sd.now$percent_of_population_fully_vaccinated, na.rm = TRUE))*100
```

- 67.28%

We can also look at the 6-number summary 

```{r}
summary(sd.now$percent_of_population_fully_vaccinated)
```

>Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?

```{r}
hist(sd.now$percent_of_population_fully_vaccinated)
```

or 

```{r}
library(ggplot2)

ggplot(sd.now) +
  aes(percent_of_population_fully_vaccinated)+
  geom_histogram(bins=15)
```


#Focus on UCSD/La Jolla

UC San Diego resides in the 92037 ZIP code area and is listed with an age 5+ population size of 36,144.

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

>Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
library(ggplot2)
ggplot(ucsd) +
  aes(as_of_date,percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```


#Comparing 92037 to other similar sized areas?

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")

head(vax.36)
```

>Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
avg.full.vax <- mean(vax.36$percent_of_population_fully_vaccinated, na.rm=TRUE)

library(ggplot2)
ggplot(ucsd) +
  aes(as_of_date,percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated") +
  geom_hline(yintercept = 0.6727567, linetype="dashed", col = "red")
 
```

>Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

>Q18. Using ggplot generate a histogram of this data.

```{r}
library(ggplot2)

ggplot(vax.36) +
  aes(percent_of_population_fully_vaccinated)+
  geom_histogram(bins=15)
```


>Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```


- Zip Code 92109 is above average at 68.78% vax where as zip code 92040 is below average at 52.05% vax. 

>Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rate Across California",
       subtitle="Only areas with a population above 36k are shown") +
  geom_hline(yintercept = 0.6727567, linetype="dashed")
```


>Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?







